---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# c14

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/c14)](https://CRAN.R-project.org/package=c14)
[![R build status](https://github.com/joeroe/c14/workflows/R-CMD-check/badge.svg)](https://github.com/joeroe/c14/actions)
[![Codecov test coverage](https://codecov.io/gh/joeroe/c14/branch/master/graph/badge.svg)](https://codecov.io/gh/joeroe/c14?branch=master)
<!-- badges: end -->

**c14** provides basic classes and functions for radiocarbon data in R.
It makes it easier to combine methods from several existing packages (e.g. rcarbon, Bchron, oxcAAR, c14bazAAR, ArchaeoPhases, stratigraphr) together and work with them in a tidy data workflow.

It was forked from [stratigraphr](https://github.com/joeroe/stratigraphr) v0.3.0.

## Installation

You can install the development version of c14 from GitHub:

```r
# install.packages("devtools")
devtools::install_github("joeroe/c14")
```

## Usage

Tidy radiocarbon calibration:

```{r eg-calibrate, message=FALSE}
library("c14")
library("dplyr")
ppnd %>% 
  filter(site == "Ganj Dareh") %>% 
  mutate(cal = c14_calibrate(cra, error, verbose = FALSE))
```

The result is a `cal` vector, which fits nicely into a data frame or tibble and comes with print methods etc.:

```{r eg-cal}
c14_calibrate(10000, 30, verbose = FALSE)
```