---
title: "Tidy radiocarbon data"
output: rmarkdown::html_vignette
bibliography: '`r system.file("REFERENCES.bib", package="c14")`'
vignette: >
  %\VignetteIndexEntry{Tidy radiocarbon data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

**c14** makes it easy and intuitive to incorporate radiocarbon dating into tidy data analysis workflows in R. 
It provides classes and functions for radiocarbon data that fit nicely in tables and work well with pipes and [dplyr](https://dplyr.tidyverse.org/) verbs.

This vignette introduces functions for the most common tasks when working with radiocarbon data, including those for calibrating, summarising, aggregating and plotting radiocarbon dates.

```{r setup, message = FALSE}
library("c14")
library("dplyr", warn.conflicts = FALSE)
library("ggplot2")
```

## Calibrating radiocarbon dates

`c14_calibrate()` calibrates conventional radiocarbon ages and their errors to a calendar timescale:

```{r calibrate-single}
c14_calibrate(5000, 30)
```

By default, `c14_calibrate()` uses the IntCal20 calibration curve [@Reimer2020], which is recommended for terrestrial dates from the northern hemisphere.
You can change this with the `c14_curve` argument (see `?IntCal20` for the available curves).

`c14_calibrate()` returns a *year with an era* (a `era::yr` vector), indicating the age is expressed as calibrated years Before Present ('cal BP').
Most functions in the package return yrs and work with them internally to ensure that calendric calculations are consistent.

Calibration is vectorised, so you can calibrate many dates at once:

```{r calibrate-vector}
ages <- c(5000, 6000, 7000)
errors <- c(30, 40, 50)

c14_calibrate(ages, errors)
```

This unlocks the real utility of the c14 package, which is embedding calibration in a tidy workflow.
For example, the `shub1_c14` dataset (provided with the package) contains 27 radiocarbon dates from the Epipalaeolithic site of Shubayqa 1 in eastern Jordan.
We can calibrate these dates within the data frame using dplyr's `mutate()` verb:

```{r calibrate-pipe}
shub1_c14 |>
  mutate(cal = c14_calibrate(c14_age, c14_error))
```

The `cal` column preserves the original data alongside the calibrated dates, so you can continue working with the full dataset.

## Summarising calibrated dates

The table in the last example showed the highest density interval of the calibrated radiocarbon date (at 95.4% probability -- sometimes called the '2-sigma' range) of the calibrated radiocarbon date, rather than its full probability distribution.
^[In fact the object returned by `c14_calibrate()` does not store the probability distribution at all, only generates it when needed.]
We can calculate this explicitly with `cal_hdi()`, which also allows us to choose another probability value:

```{r cal_hdi}
shub1_c14 |>
  mutate(
    cal = c14_calibrate(c14_age, c14_error),
    cal_hdi_1sigma = cal_hdi(cal, interval = 0.683)
  ) |>
  select(lab_id, c14_age, c14_error, cal, cal_hdi_1sigma)
```

`cal_hdr()` always returns a single region of the probability distribution – the shortest interval containing the specified probability mass.
Calibrated radiocarbon dates are often multimodal, so it can be more accurate to consider one or more highest density *regions*, which we can calculate with `cal_hdr()`:

```{r cal_hdr}
c14_calibrate(8800, 30) |>
  cal_hdr()
```

The inverse of these two functions is `cal_probability()`, which calculates the probability that a calibrated date falls within a specified time interval:

```{r cal_probability}
x <- c14_calibrate(1116, 30)
cal_probability(x, 970, 1060) # cal BP
```

Other functions for summarising calibrated radiocarbon dates include various methods for calculating a point estimate (see `?cal_point`).
Generally speaking, there is no 'good' point estimate of a calibrated radiocarbon distribution [@Michczynski2007], but these can nevertheless be useful when working with a range or the full probability distribution is not an option.
In this case, the mode is a reasonable choice:

```{r cal_mode}
shub1_c14 |>
  mutate(
    cal = c14_calibrate(c14_age, c14_error),
    cal_mode = cal_mode(cal)
  ) |>
  select(lab_id, c14_age, c14_error, cal, cal_mode)
```

Note that here we mean `summarise` in the sense of deriving summary statistics from the calendar probability distribution.
They return a vector of the same length as the input and therefore in dplyr terms they are *mutating* functions, not *summarising* functions;
operations that can be used with `dplyr::summarise()` are covered in the section below on [aggregating calibrated dates](#aggregating-calibrated-dates).

## Plotting calibrated dates

### Base R plots

The package provides a default `plot()` method for calibrated dates which shows their probability distributions and the calibration curve.

```{r plot}
shub1_cal <- c14_calibrate(shub1_c14$c14_age, shub1_c14$c14_error)
plot(shub1_cal[1])
```

This default method plots all the calibrated dates passed to it.
^[By default `plot(cal)` only draws up to 20 plots at a time; this can be controlled with `max.plot`.]
You can arrange multiple plots with, for example, `par(mfrow)`:

```{r plot-multi, fig.height = 10}
par(mfrow = c(2, 1))
plot(shub1_cal[1:2])
par(mfrow = c(1, 1))
```

### ggplot2

Because c14 works with standard R data structures, it is straightforward to build custom ggplots.
The key is to extract the probability distribution data and tidy it into a data frame.

The `cal_dist()` function returns a list of data frames, each with `age` and `pdens` columns.
We can combine these into a single tidy data frame for plotting:

```{r ggplot-setup}
tidy_cal <- function(cal_vec) {
  dists <- c14:::cal_dist(cal_vec)
  purrr::imap_dfr(dists, function(df, i) {
    df$label <- paste0("Date ", i)
    df
  })
}
```

```{r ggplot-single, fig.width = 6, fig.height = 4}
dist_df <- tidy_cal(shub1_cal[1])
dist_df$label <- shub1_c14$lab_id[1]

ggplot(dist_df, aes(x = age, y = pdens)) +
  geom_ribbon(aes(ymin = 0, ymax = pdens), fill = "steelblue", alpha = 0.3) +
  geom_line(colour = "steelblue") +
  labs(x = "Calendar age (cal BP)", y = "Probability density") +
  scale_x_reverse() +
  theme_minimal()
```

You can plot multiple dates by faceting.
Here we add the lab ID as a grouping column before combining the data frames:

```{r ggplot-facet, fig.width = 7, fig.height = 8}
dist_df <- tidy_cal(shub1_cal[1:6])
n_per_date <- sapply(c14:::cal_dist(shub1_cal[1:6]), nrow)
dist_df$lab_id <- rep(shub1_c14$lab_id[1:6], times = n_per_date)

ggplot(dist_df, aes(x = age, y = pdens)) +
  geom_ribbon(aes(ymin = 0, ymax = pdens), fill = "steelblue", alpha = 0.3) +
  geom_line(colour = "steelblue") +
  facet_wrap(~lab_id, ncol = 2, scales = "free_y") +
  labs(x = "Calendar age (cal BP)", y = "Probability density") +
  scale_x_reverse() +
  theme_minimal()
```

This makes it easy to compare dates across sites, phases, or any other grouping in your data.

## Aggregating calibrated dates

c14 provides two methods for combining multiple calibrated dates into a single summary distribution.

### Summed probability distributions

`cal_sum()` adds up the probability distributions of all dates in a vector.
This is the simplest aggregation method.

```{r cal-sum}
shub1_cal <- c14_calibrate(shub1_c14$c14_age, shub1_c14$c14_error)

summed <- cal_sum(shub1_cal)
```

```{r cal-sum-plot, fig.width = 7, fig.height = 4}
sum_df <- vctrs::vec_data(summed)[[1]]

ggplot(sum_df, aes(x = age, y = pdens)) +
  geom_ribbon(aes(ymin = 0, ymax = pdens), fill = "darkred", alpha = 0.3) +
  geom_line(colour = "darkred") +
  labs(x = "Calendar age (cal BP)", y = "Summed probability") +
  scale_x_reverse() +
  theme_minimal()
```

You can also use the `sum()` method:

```{r cal-sum-method}
sum(shub1_cal)
```

### Composite kernel density estimation

`cal_density()` uses composite kernel density estimation (cKDE) to produce a smoother summary distribution.
It resamples from each date's probability distribution and fits a kernel density estimate, with optional bootstrapping to estimate uncertainty.

```{r cal-density}
density_est <- cal_density(shub1_cal, bw = 30, times = 10)
density_est
```

```{r cal-density-plot, fig.width = 7, fig.height = 4}
ggplot(density_est, aes(x = age, y = .estimate)) +
  geom_ribbon(aes(ymin = .estimate - .error, ymax = .estimate + .error),
              fill = "darkred", alpha = 0.3) +
  geom_line(colour = "darkred") +
  labs(x = "Calendar age (cal BP)", y = "Density") +
  scale_x_reverse() +
  theme_minimal()
```

The bootstrapped standard error (`.error`) gives a measure of uncertainty in the density estimate.

### Stratified aggregation

Both methods can be stratified by a grouping variable.
This is useful for comparing periods or phases.

To compare groups, run the aggregation separately for each stratum:

```{r cal-density-strata, fig.width = 7, fig.height = 5}
phases <- unique(shub1_c14$phase)

phase_densities <- purrr::map(phases, function(p) {
  idx <- shub1_c14$phase == p
  cal_density(shub1_cal[idx], bw = 30, times = 10) |>
    mutate(phase = p)
}) |> bind_rows()

ggplot(phase_densities, aes(x = age, y = .estimate)) +
  geom_ribbon(aes(ymin = .estimate - .error, ymax = .estimate + .error),
              fill = "steelblue", alpha = 0.3) +
  geom_line(colour = "steelblue") +
  facet_wrap(~phase, ncol = 1) +
  labs(x = "Calendar age (cal BP)", y = "Density") +
  scale_x_reverse() +
  theme_minimal()
```

## References
